MYC overrides HIF-1α to regulate proliferating primary cell metabolism in hypoxia

### 2023-04-04 manuscript/manuscript.qmd
draft figure 3

The goal of this study was to characterize hypoxia-induced metabolic changes in proliferating primary LFs and PASMCs. To accomplish this goal, we adopted a metabolic flux analysis technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates. Metabolic flux analysis fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a computational model of cell metabolism [@antoniewicz2018]. This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, through intracellular metabolic pathways, and into cell biomass and metabolic by-products [@young2014]. These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015]. Exponential growth phase is thought to reflect metabolic pseudo-steady state as cells in culture steadily divide at their maximal condition-specific rate, provided nutrient supply does not become limiting [@buescher2015; @ahn2011]. Thus, we first set out to define experimental conditions to capture exponential growth phase in normoxic and hypoxic cultures.

Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program (**Figure 1A**). We selected 0.5% oxygen for hypoxia as this level yielded the most reproducible phenotypic differences compared to 21% oxygen culture, while being physiologically relevant and above the K~M~ of cytochrome *c* oxidase (electron transport chain complex IV) for oxygen [@lee2020; @wenger2015]. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure 1B**). LFs cultured in 0.5% oxygen grew more slowly than LFs cultured in 21% oxygen (**Figure 1C**), but slower growth was not associated with decreased cell viability (**Figure 1 - figure supplement 1**). As anticipated, hypoxic cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figure 2**). These changes persisted for the duration of the experimental time course.

>>>>

The goal of this study was to characterize hypoxia-induced metabolic changes in proliferating primary LFs and PASMCs. To accomplish this goal, we adopted a metabolic flux analysis technique that enabled us to link intracellular metabolic fluxes to cell proliferation rates. Metabolic flux analysis fits cell proliferation rate, extracellular flux measurements, and ^13^C intracellular isotope labeling patterns to a computational model of cell metabolism [@antoniewicz2018]. This analysis reconstructs comprehensive flux maps that depict the flow of carbon from extracellular substrates, through intracellular metabolic pathways, and into cell biomass and metabolic by-products [@young2014]. These models assume that cells are at a metabolic pseudo-steady state over the experimental time course [@buescher2015]. Exponential growth phase is thought to reflect metabolic pseudo-steady state as cells in culture steadily divide at their maximal condition-specific rate, provided nutrient supply does not become limiting [@buescher2015; @ahn2011]. Thus, we first set out to define experimental conditions to capture the exponential growth phase in normoxic and hypoxic cultures.

Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program (**Figure 1A**). We selected 0.5% oxygen for hypoxia as this level yielded the most reproducible phenotypic differences compared to 21% oxygen, while being physiologically relevant and above the K~M~ of cytochrome *c* oxidase (electron transport chain complex IV) for oxygen [@lee2020; @wenger2015]. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure 1B**). LFs cultured in 0.5% oxygen grew more slowly than LFs cultured in 21% oxygen (**Figure 1C**), but slower growth was not associated with decreased cell viability (**Figure 1 - figure supplement 1**). As anticipated, hypoxic cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figure 2**). These changes persisted for the duration of the experimental time course.

Having identified experimental conditions for exponential growth, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figure 3**). Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, metabolite degradation rates, and medium evaporation rate [@murphy2013]. Interestingly, while we observed a modest increase in glucose uptake, we found that hypoxia actually decreased lactate efflux (**Figure 3A**). This finding was confirmed by measuring the rate of [U-^13^C~3~]-lactate efflux from LFs cultured in the presence of [U-^13^C~6~]-glucose (**Figure 3 - figure supplement 1**).

This decrease in lactate efflux occurred despite activation of the HIF-1α transcriptional program as reflected by increased expression of GLUT1 and LDHA. To test if more severe hypoxia would augment glycolysis, we cultured cells in 0.2% ambient oxygen (**Fig S2**). Under these conditions, we observed no change in glucose or lactate fluxes, similar to 0.5% oxygen culture. To test if this unexpected response was unique to LFs, we next studied PASMCs under 0.5% oxygen conditions (**Fig S3**). PASMCs grew faster than LFs, and so samples were collected every 12 h for 48 h for these cells. Again, similar to LFs at 0.5% and 0.2% oxygen, we observed no change in glucose uptake and reduced lactate efflux in PASMCs regardless of HIF-1α stabilization. Together, these data suggest that hypoxia uncouples HIF target gene expression and glycolytic flux in proliferating primary cells.
